Cleaning Script

### 2025-04-21 code/eda-code/eda.qmd
Color Corrections

```{r}
#load needed packages. make sure they are installed
library(here) 
library(dplyr)
library(ggplot2)
library(zoo)
library(patchwork)
library(e1071)
library(scales)
library(RColorBrewer)
library(corrplot)
library(reshape2)

@@

case_data <- readRDS(here("data", "processed-data", "processedcasedata.rds"))

data<- data %>%
   mutate(incidence_rate = if_else(incidence_rate < 0, 0, incidence_rate))
```

DataVisualization

@@

From the graphs, kurtosis test, and skewness test, it is apparent that daily new case incidence is not normally distribued and heavily skewed towards 0 and other lower values. The second graph, new cases over 500, was graphed to show how certain days do reach high case incidence.
```{r}

density_cases <- data %>%
  ggplot( aes(x=new_cases)) +
    geom_density(fill="#69b3a2", color="#e9ecef", alpha=0.8) +
  labs(title = "Distribution of Daily New Case Incidence", 
       x = "Daily New Cases",
       y = "Density") +
  theme_bw()

print(density_cases)

figure_file = here("results","figures","density_cases.png")
ggsave(filename = figure_file, plot=density_cases)

density_cases_500 <- data %>%
  filter(new_cases > 500) %>%
  ggplot( aes(x=new_cases)) +
    geom_density(fill="#69b3a2", color="#e9ecef", alpha=0.8) +
  labs(title = "Distribution of Daily New Case Incidence (Over 500)", 
       x = "Daily New Cases",
       y = "Density") +
  theme_bw()

print(density_cases_500)

figure_file = here("results","figures","density_cases_500.png")
ggsave(filename = figure_file, plot=density_cases_500)

print(kurtosis(data$new_cases))
print(skewness(data$new_cases))
```
A histogram of population density was created. I hypothesize that population density plays a large role in case incidence. Population density is also skewed, however, there are some notable counties that do have large populationd densities.
```{r}
data_summary <- data %>%
  group_by(county) %>%
  summarize(pop_density = mean(pop_density, na.rm = TRUE))

histo_pop_density <- ggplot(data_summary, aes(x = pop_density)) + 
  geom_histogram(bins = 30, fill = "#69b3a2", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Population Density (2-Year Average per County)", 
       x = "Population Density (Calculated as Population Count / km squared)", y = "Frequency") + theme_bw()

print(histo_pop_density)

figure_file = here("results","figures","histo_pop_density.png")
ggsave(filename = figure_file, plot=histo_pop_density)
```

@@

# Create dataframe of average mobility per county per date
ga_average_mobility <- data %>%
  group_by(date) %>%
  summarise(
    retail_mean = mean(retail_and_recreation_percent_change_from_baseline, na.rm = TRUE),
    grocery_mean = mean(grocery_and_pharmacy_percent_change_from_baseline, na.rm = TRUE),
    workplaces_mean = mean(workplaces_percent_change_from_baseline, na.rm = TRUE),
    residential_mean = mean(residential_percent_change_from_baseline, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  arrange(date) %>%  # Ensure dates are sorted
  mutate(
    retail_mean_7day = rollmean(retail_mean, k = 7, fill = NA, align = "right"),
    grocery_mean_7day = rollmean(grocery_mean, k = 7, fill = NA, align = "right"),
    workplaces_mean_7day = rollmean(workplaces_mean, k = 7, fill = NA, align = "right"),
    residential_mean_7day = rollmean(residential_mean, k = 7, fill = NA, align = "right")
  )

# Plot the 7-day averages for each mobility type
mobility_graph <- ggplot(ga_average_mobility, aes(x = date)) +
  geom_line(aes(y = grocery_mean_7day, color = "Grocery & Pharmacy")) +
  geom_line(aes(y = workplaces_mean_7day, color = "Workplaces")) +
  geom_line(aes(y = residential_mean_7day, color = "Residential")) +
  geom_line(aes(y = retail_mean_7day, color = "Retail")) +
  labs(
    title = "7-Day Moving Average of Percent Change in Mobility in Georgia Over Time",
    x = "Date",
    y = "7-Day Average of Percent Change in Mobility",
    color = "Mobility Type"
  ) +
  theme_bw()

print(mobility_graph)

@@

```{r}
# Create one plot per mobility type
p1 <- ggplot(ga_average_mobility, aes(x = date, y = grocery_mean_7day)) +
  geom_line(color = "steelblue") +
  labs(title = "Grocery & Pharmacy", x = "Date", y = "7-Day Avg % Change") +
  theme_bw() +
  coord_cartesian(ylim = c(-50, 20))

p2 <- ggplot(ga_average_mobility, aes(x = date, y = workplaces_mean_7day)) +
  geom_line(color = "darkred") +
  labs(title = "Workplaces", x = "Date", y = "7-Day Avg % Change") +
  theme_bw() +
  coord_cartesian(ylim = c(-50, 20))

p3 <- ggplot(ga_average_mobility, aes(x = date, y = residential_mean_7day)) +
  geom_line(color = "forestgreen") +
  labs(title = "Residential", x = "Date", y = "7-Day Avg % Change") +
  theme_bw() +
  coord_cartesian(ylim = c(-50, 20))

p4 <- ggplot(ga_average_mobility, aes(x = date, y = retail_mean_7day)) +
  geom_line(color = "purple") +
  labs(title = "Retail", x = "Date", y = "7-Day Avg % Change") +
  theme_bw() +
  coord_cartesian(ylim = c(-50, 20))

# Arrange in a 2x2 grid
mobility_grid <- (p1 | p2) / (p3 | p4)

# Show it
print(mobility_grid)

@@

first_date <- min(data$date)
last_date <- max(data$date)

cases_graph <- case_data %>%
  group_by(date) %>%
  summarise(total_cases = sum(total_cases, na.rm = TRUE)) %>%
  arrange(date) %>%
  ggplot(aes(x = date, y = total_cases)) +
    geom_area(fill = "#69b3a2", alpha = 0.5) +
    geom_line(color = "#69b3a2") +
    ylab("Total Cases (in GA)") +
    scale_y_continuous(labels = label_number()) +
   scale_x_date(
      breaks = seq(from = first_date, to = last_date, by = "6 months"),  # Breaks every 3 months
      labels = date_format("%b %Y")  # Format labels as month year (e.g., Jan 2020)
    ) +
    theme_bw() +
  labs(title = "Total Cases (in GA) Over Time",
       x = "Date (2020 - 2022)",
       y = "Total Cases")

print(cases_graph)

figure_file = here("results","figures","cases_graph.png")
ggsave(filename = figure_file, plot=cases_graph)

incidence_graph <- data %>%
  group_by(date) %>%
  summarise(new_cases = sum(new_cases, na.rm = TRUE)) %>%
  arrange(date) %>%
  ggplot(aes(x = date, y = new_cases)) +
    geom_area(fill = "#69b3a2", alpha = 0.5) +
    geom_line(color = "#69b3a2") +
    ylab("Daily New Cases") +
     scale_x_date(
      date_breaks = "3 months",  # Automatically set the breaks to every 3 months
      date_labels = "%b %Y"      # Format labels as month-year (e.g., Jan 2020)
    ) +
    theme_bw() +
  labs(title = "Daily New Cases (in GA) Over Time",
       x = "Date (2020 - 2022)",
       y = "Total Cases")

print(incidence_graph)

figure_file = here("results","figures","incidence_graph.png")
ggsave(filename = figure_file, plot=incidence_graph)
```

This code chunk plots a scatter plot of a county's population count against its final total cases. 
```{r}

year_end_cases <- data %>%
  mutate(year = year(date)) %>%
  group_by(county, year) %>%
  slice_max(order_by = date, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(county, year, total_cases)

pop_density_info <- data %>%
  mutate(year = year(date)) %>%
  select(county, year, pop_density) %>%
  distinct()

plot_df <- left_join(year_end_cases, pop_density_info, by = c("county", "year"))

year_end_cases_graph <- ggplot(plot_df, aes(x = pop_density, y = total_cases)) +
  geom_point(color = "#69b3a2", alpha = 0.6) +
  facet_wrap(~year) +
  labs(
    x = "Population Density",
    y = "Total COVID Cases (End of Year)",
    title = "Year-End COVID Cases vs. Population Density by County"
  ) +
  theme_bw()

print(year_end_cases_graph)

figure_file = here("results","figures","year_end_cases_graph.png")
ggsave(filename = figure_file, plot=year_end_cases_graph)
```
Correlation Analysis

This code chunk entails correlation testing through Spearman's and the creation of a matrix. First, lagged variables for case count was created for the data.
```{r}

data <- data %>%
  group_by(county) %>%
  arrange(date) %>%
  mutate(
    lag_1 = lag(new_cases, 1),
    lag_7 = lag(new_cases, 7),
    lag_14 = lag(new_cases, 14)
  ) %>%
  ungroup()

data <- data %>% drop_na()

@@

# Create plot for matrix
matrix <- ggplot(cor_matrix_melted, aes(x = Var1, y = Var2)) +
  geom_point(aes(size = abs(value), color = value)) +
  geom_text(aes(label = round(value, 2)), color = "black", size = 3) +
  scale_color_gradient2(low = "#BFA2DB", high = "#92C7CF", mid = "white",
                        midpoint = 0, limit = c(-1, 1), name = "Correlation") +
  scale_size(range = c(2, 10), guide = "none") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Correlation Heatmap (Spearman's)", x = "Variables", y = "Variables")

print(matrix)

figure_file = here("results","figures","matrix.png")
ggsave(filename = figure_file, plot=matrix)
```

>>>>

```{r}
# load needed packages
library(here) 
library(dplyr)
library(ggplot2)
library(zoo)
library(patchwork)
library(e1071)
library(scales)
library(RColorBrewer)
library(corrplot)
library(reshape2)
library(lubridate)
library(tidyr)
```

@@

# Load full case data
case_data <- readRDS(here("data", "processed-data", "processedcasedata.rds"))

# Make incidence rate non negative
data<- data %>%
   mutate(incidence_rate = if_else(incidence_rate < 0, 0, incidence_rate))
```

Data Visualization

@@

From the graphs, kurtosis test, and skewness test, it is apparent that daily new case incidence is not normally distribued and heavily skewed towards 0 and other lower values. The second graph, new cases over 500, was graphed to show how certain days do reach high case incidence.
```{r}
# Plot density of case incidence
density_cases <- ggplot(data, aes(x=new_cases)) +
    geom_density(fill="#69b3a2", color="#69b3a2", alpha=0.8) +
  labs(title = "Distribution of Daily New Case Incidence", 
       x = "Daily New Cases",
       y = "Density") +
  theme_bw()

# print
print(density_cases)

# Save
figure_file = here("results","figures","density_cases.png")
ggsave(filename = figure_file, plot=density_cases)

# plot density of case incidence over 500
density_cases_500 <- data %>%
  filter(new_cases > 500) %>%
  ggplot( aes(x=new_cases)) +
    geom_density(fill="#69b3a2", color="#69b3a2", alpha=0.8) +
  labs(title = "Distribution of Daily New Case Incidence (Over 500)", 
       x = "Daily New Cases",
       y = "Density") +
  theme_bw()

# print
print(density_cases_500)

# save
figure_file = here("results","figures","density_cases_500.png")
ggsave(filename = figure_file, plot=density_cases_500)

print(kurtosis(data$new_cases))
print(skewness(data$new_cases))
```

A histogram of population density was created. I hypothesize that population density plays a large role in case incidence. Population density is also skewed, however, there are some notable counties that do have large populationd densities.
```{r}
# average population density throughout all three years
data_summary <- data %>%
  group_by(county) %>%
  summarize(pop_density = mean(pop_density, na.rm = TRUE))

# plot summary of population density
histo_pop_density <- ggplot(data_summary, aes(x = pop_density)) + 
  geom_histogram(bins = 30, fill = "#69b3a2", color = "#69b3a2", alpha = 0.7) +
  labs(title = "Histogram of Population Density (2-Year Average per County)", 
       x = "Population Density (Calculated as Population Count / km squared)", y = "Frequency") + theme_bw()

# print
print(histo_pop_density)

# save
figure_file = here("results","figures","histo_pop_density.png")
ggsave(filename = figure_file, plot=histo_pop_density)
```

@@

# Create dataframe of average mobility per county per date
ga_average_mobility <- data %>%
  group_by(date) %>%
  summarise(
    retail_mean = mean(retail_and_recreation_percent_change_from_baseline, na.rm = TRUE),
    grocery_mean = mean(grocery_and_pharmacy_percent_change_from_baseline, na.rm = TRUE),
    workplaces_mean = mean(workplaces_percent_change_from_baseline, na.rm = TRUE),
    residential_mean = mean(residential_percent_change_from_baseline, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  arrange(date) %>%
  mutate(
    retail_mean_7day = rollmean(retail_mean, k = 7, fill = NA, align = "right"),
    grocery_mean_7day = rollmean(grocery_mean, k = 7, fill = NA, align = "right"),
    workplaces_mean_7day = rollmean(workplaces_mean, k = 7, fill = NA, align = "right"),
    residential_mean_7day = rollmean(residential_mean, k = 7, fill = NA, align = "right")
  )

# Set custom colors
custom_colors <- c(
  "Grocery & Pharmacy" = "#69b3a2",
  "Workplaces" = "#BFA2DB",
  "Residential" = "#F2C29B",
  "Retail" = "#DBA2A2"
)

# Plot the 7-day averages for each mobility type
mobility_graph <- ggplot(ga_average_mobility, aes(x = date)) +
  geom_line(aes(y = grocery_mean_7day, color = "Grocery & Pharmacy")) +
  geom_line(aes(y = workplaces_mean_7day, color = "Workplaces")) +
  geom_line(aes(y = residential_mean_7day, color = "Residential")) +
  geom_line(aes(y = retail_mean_7day, color = "Retail")) +
  scale_color_manual(values = custom_colors) +
  labs(
    title = "7-Day Moving Average of Percent Change in Mobility in Georgia Over Time",
    x = "Date",
    y = "7-Day Average of Percent Change in Mobility",
    color = "Mobility Type"
  ) +
  theme_bw()

# print
print(mobility_graph)

@@

```{r}
# Create one plot per mobility type
grocery <- ggplot(ga_average_mobility, aes(x = date, y = grocery_mean_7day)) +
  geom_line(color = "#69b3a2") +
  labs(title = "Grocery & Pharmacy", x = "Date", y = "7-Day Avg % Change") +
  theme_bw() +
  coord_cartesian(ylim = c(-50, 20))

workplace <- ggplot(ga_average_mobility, aes(x = date, y = workplaces_mean_7day)) +
  geom_line(color = "#BFA2DB") +
  labs(title = "Workplaces", x = "Date", y = "7-Day Avg % Change") +
  theme_bw() +
  coord_cartesian(ylim = c(-50, 20))

resident <- ggplot(ga_average_mobility, aes(x = date, y = residential_mean_7day)) +
  geom_line(color = "#F2C29B") +
  labs(title = "Residential", x = "Date", y = "7-Day Avg % Change") +
  theme_bw() +
  coord_cartesian(ylim = c(-50, 20))

retail <- ggplot(ga_average_mobility, aes(x = date, y = retail_mean_7day)) +
  geom_line(color = "#DBA2A2") +
  labs(title = "Retail", x = "Date", y = "7-Day Avg % Change") +
  theme_bw() +
  coord_cartesian(ylim = c(-50, 20))

# Combine in 2x2
mobility_grid <- (grocery | workplace) / (resident | retail)

# print for visual inspection
print(mobility_grid)

@@

# Get first and last date of data
first_date <- min(data$date)
last_date <- max(data$date)

# Create graph of total case count through the 3 years
cases_graph <- case_data %>%
  group_by(date) %>%
  summarise(total_cases = sum(total_cases, na.rm = TRUE)) %>%
  arrange(date) %>%
  ggplot(aes(x = date, y = total_cases)) +
    geom_area(fill = "#69b3a2", alpha = 0.5) +
    geom_line(color = "#69b3a2") +
    ylab("Total Cases (in GA)") +
    scale_y_continuous(labels = label_number()) +
   scale_x_date(
      breaks = seq(from = first_date, to = last_date, by = "6 months"),
      labels = date_format("%b %Y")
    ) +
    theme_bw() +
  labs(title = "Total Cases (in GA) Over Time",
       x = "Date (2020 - 2022)",
       y = "Total Cases")

# print
print(cases_graph)

# save
figure_file = here("results","figures","cases_graph.png")
ggsave(filename = figure_file, plot=cases_graph)

# plot incidence
incidence_graph <- data %>%
  group_by(date) %>%
  summarise(new_cases = sum(new_cases, na.rm = TRUE)) %>%
  arrange(date) %>%
  ggplot(aes(x = date, y = new_cases)) +
    geom_area(fill = "#69b3a2", alpha = 0.5) +
    geom_line(color = "#69b3a2") +
    ylab("Daily New Cases") +
     scale_x_date(
      date_breaks = "3 months",
      date_labels = "%b %Y"
    ) +
    theme_bw() +
  labs(title = "Daily New Cases (in GA) Over Time",
       x = "Date (2020 - 2022)",
       y = "Total Cases")

# print
print(incidence_graph)

# save
figure_file = here("results","figures","incidence_graph.png")
ggsave(filename = figure_file, plot=incidence_graph)
```

This code chunk plots a scatter plot of a county's population count against its final total cases. 
```{r}
# create df with total case count at the end of the year
year_end_cases <- data %>%
  mutate(year = year(date)) %>%
  group_by(county, year) %>%
  slice_max(order_by = date, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(county, year, total_cases)

# pick population density of the years
pop_density_info <- data %>%
  mutate(year = year(date)) %>%
  select(county, year, pop_density) %>%
  distinct()

# combine dfs for ploting
plot_df <- left_join(year_end_cases, pop_density_info, by = c("county", "year"))

# plot
year_end_cases_graph <- ggplot(plot_df, aes(x = pop_density, y = total_cases)) +
  geom_point(color = "#69b3a2", alpha = 0.6) +
  facet_wrap(~year) +
  labs(
    x = "Population Density",
    y = "Total COVID Cases (End of Year)",
    title = "Year-End COVID Cases vs. Population Density by County"
  ) +
  theme_bw()

# print
print(year_end_cases_graph)

# save
figure_file = here("results","figures","year_end_cases_graph.png")
ggsave(filename = figure_file, plot=year_end_cases_graph)
```
Correlation Analysis

This code chunk entails correlation testing through Spearman's and the creation of a matrix. First, lagged variables for case count was created for the data.
```{r}
# create lagged variables
data <- data %>%
  group_by(county) %>%
  arrange(date) %>%
  mutate(
    lag_1 = lag(new_cases, 1),
    lag_7 = lag(new_cases, 7),
    lag_14 = lag(new_cases, 14)
  ) %>%
  ungroup()

# Drop NAs made as a result
data <- data %>% drop_na()

@@

# Create plot for matrix
matrix <- ggplot(cor_matrix_melted, aes(x = Var1, y = Var2)) +
  geom_point(aes(size = abs(value), color = value)) +
  geom_text(aes(label = round(value, 2)), color = "black", size = 3) +
  scale_color_gradient2(low = "#BFA2DB", high = "#69b3a2", mid = "white",
                        midpoint = 0, limit = c(-1, 1), name = "Correlation") +
  scale_size(range = c(2, 10), guide = "none") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Correlation Heatmap (Spearman's)", x = "Variables", y = "Variables")

# Print
print(matrix)

# save
figure_file = here("results","figures","matrix.png")
ggsave(filename = figure_file, plot=matrix)
```

### 2025-04-21 code/processing-code/processing-file.qmd
Color Corrections

```{r}
# load needed packages
library(here) 
library(dplyr)
library(tidyr)
library(lubridate)
library(readxl)
library(stringr)
library(skimr)

library(tigris)
library(sf)
```

>>>>

```{r}
# load needed packages
library(here) 
library(dplyr)
library(tidyr)
library(lubridate)
library(readxl)
library(stringr)
library(skimr)
library(tigris)
library(sf)
```
